---
title: "Gene Set Phenotypic Profile Comparison"
author: "Hannah Blau and Peter Robinson"
date: "6/27/2018"
output: html_notebook
params:
  to.compare: "ENTREZ_gene_sets.tsv"
  gene.info: "human_protein_coding_genes.tsv"
  pheno.to.gene: "ALL_SOURCES_ALL_FREQUENCIES_phenotype_to_genes.txt"
  num.iter: 10000
---
This notebook compares two disjoint sets of genes to see whether their phenotypic profiles are more similar than one would expect by chance.

### Inputs
* file containing list of ENTREZ gene ids for each gene set (ids are tab-separated, blank line marks the end of the first set). The gene sets may be of different sizes.
* file containing gene id and gene name for all human protein coding genes, extracted from [ftp.ncbi.nih.gov/gene/DATA/gene_info.gz](http://ftp.ncbi.nih.gov/gene/DATA/gene_info.gz) with
```{bash, eval = FALSE}
zgrep ^9606 gene_info.gz  | grep protein-coding | cut -f2,3 > human_protein_coding_genes.tsv
```
* file mapping HPO phenotype id/name to gene id/name: [ALL_SOURCES_ALL_FREQUENCIES_phenotype_to_genes.txt](http://compbio.charite.de/jenkins/job/hpo.annotations.monthly/lastSuccessfulBuild/artifact/annotation/ALL_SOURCES_ALL_FREQUENCIES_phenotype_to_genes.txt)
* number of iterations in loop to estimate p-value through simulation ($n$)

### Algorithm
1. For each set of genes ($G_{1}$, $G_{2}$), create a set containing all phenotypes related to those genes ($P_{1}$, $P_{2}$).
2. Compute Jaccard similarity of $P_{1}$ and $P_{2}$.

To estimate the p value by simulation, hold $G_{1}$ fixed and choose the second gene set at random from all human protein-coding genes.

3. Iterate $n$ times:
+ Select from all human protein-coding genes a set $G_{3}$ equal in size to $G_{2}$.
+ Create the corresponding set of phenotypes $P_{3}$ and compute the Jaccard similarity of $P_{1}$ and $P_{3}$. Note that $P_{3}$ might be empty.
+ Maintain running count $m$ of how many times the similarity of $P_{1}$ and $P_{3}$ is $\geq$ similarity of $P_{1}$ and $P_{2}$.

4. Calculate p value: $m$ / $n$

### Load library
Meyer D and Hornik K (2009). “Generalized and Customizable Sets in R.” _Journal of Statistical Software_, 31(2), pp. 1–27. doi: 10.18637/jss.v031.i02.0001459
```{r}
library(sets, quietly = TRUE)
```
### Read input files
```{r}
human.genes <- read.table(params$gene.info, sep = "\t", skip = 1,
                          col.names = c("GeneID", "Symbol"), colClasses = c("integer", "character"))
pheno.2.gene <- read.delim(params$pheno.to.gene, header = FALSE, sep = "\t", skip = 1,
                           col.names = c("HPO.ID", "HPO.Name", "Gene.ID", "Gene.Name"),
                           colClasses = c("character", "character", "integer", "character"))
genes.file <- file(params$to.compare, "r")
lines.read <- readLines(genes.file)
close(genes.file)
```
### Create gene sets
Create two gene sets, one for GPI pathway genes and the other for genes that code the GPI anchored proteins. Each is a set of ENTREZ gene ids, treated as integers. Parse the strings read from __params$to.compare__ to create sets of integers. A blank line indicates the end of the first set and the beginning of the second set. The set constructor expects a vector, not a list, of elements.
```{r}
collector <- c()
gene.sets <- list()
for (i in 2:length(lines.read)) {
  if (lines.read[i] == c("")) {
    gene.sets[[1]] <- as.set(unlist(lapply(collector, as.integer)))
    collector <- c()
  }
  else {
    collector <- c(collector, unlist(strsplit(lines.read[i], "\t")))
  }
}
gene.sets[[2]] <- as.set(unlist(lapply(collector, as.integer)))
```
### Compute Jaccard similarity of phenotype sets
For each set of genes, we want the corresponding set of phenotypes that includes every phenotype reported for any of the genes. First, define a function that takes a single gene id as its argument and returns the set of phenotypes associated with that gene.
```{r}
find.phenotypes <- function(gene.id) {
  phenotypes <- subset(pheno.2.gene, select=HPO.ID, subset=(Gene.ID==gene.id))
  as.set(phenotypes$HPO.ID)
}
```
Next, define a function that takes a set of gene ids as its argument and returns the set of all phenotypes associated with those genes.
```{r}
findall.phenotypes <- function(gene.ids) {
  phenotypes <- as.set(c())       # initialize phenotypes to empty set
  for (gene.id in gene.ids) {
    phenotypes <- set_union(phenotypes, find.phenotypes(gene.id))
  }
  phenotypes
}
```
Now, use this function to find the sets of phenotypes $P_{1}$ and $P_{2}$ for the two input sets of genes $G_{1}$ and $G_{2}$. Calculate the Jaccard similarity of $P_{1}$ and $P_{2}$.
```{r}
p1 <- findall.phenotypes(gene.sets[[1]])
p2 <- findall.phenotypes(gene.sets[[2]])
p1p2.similarity <- set_similarity(p1, p2, method = "Jaccard")
```

### Random sampling from human protein coding genes
Repeat __params\$num.iter__ times: choose at random a set $G_{3}$ of genes from __params\$gene.info__ equal in size to $G_{2}$. Find the set of associated phenotypes $P_{3}$ and calculate the Jaccard similarity with $P_{1}$. Count how many times $P_{1}$'s similarity to $P_{3}$ is $\geq$ its similarity to $P_{2}$.

```{r}
m <- 0
cardinality.g2 <- length(gene.sets[[2]])
for (i in 1:params$num.iter) {
  g3 <- sample(human.genes$GeneID, cardinality.g2)
  p3 <- findall.phenotypes(g3)
  if (set_similarity(p1, p3, method = "Jaccard") >= p1p2.similarity) {
    m <- m + 1
  }
}
```
### Calculate p value
P value after `r params$num.iter` iterations.
```{r}
options(digits = 4)
cat("p value is : ", m/params$num.iter, "\n")
```
